Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-26 paper/paper.tex
Fix the typos pointed out by Cassia

Tutorial material on logical relations~\citep{skorstengaard2019introduction,
  harpertait, harperkripke, pierce2002types,
  pierce2004advanced,harper2016practical} is primarily focused on systems
with simple or polymorphic types. In that context, syntactic logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation. However, neither of these techniques can be used to define a
logical relation in the context of a predicative dependent type theory, so a
young researcher might be excused for thinking that proofs that use logical
relations are not applicable for such languages.

But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs for dependent type theories}
Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code. As a result, their use of logical relations arguments
difficult to isolate from their surrounding contexts, and thus, not accessible
to casual readers.

@@

Not only does this connection aid readers that wish to, like us, adopt proof
assistants for their day-to-day use. We also find that this precision is
important for conveying the proof technique itself. Unlike properties that are
derivable through syntactic means, proofs by logical relation make demands on
the strength of the logic in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying
metatheory % may struggle to even
% convey the validity of their definition and
requires substantial
effort from the reader to even understand whether it is theoretically
possible to encode such proofs in a given ambient logic, such as those found
in modern proof assistants.

% The close correspondence between a typeset proof and a mechanized
% proof comes at the potential risk that the artifacts specific to
% theorem provers may become visible in the presented proof.
%However, for proofs by logical relation for
%dependent types, we find our approach advantageous for a few reasons.
Furthermore, while a proof assistant may reject perfectly valid definitions,
it helps us keep our definition precise and unambiguous. A precise definition
is crucial as the reasoning principles we are allowed to use is directly tied
to the definition, which is sometimes obscured by concise notations that
diverge too far from a formal proof.

@@

In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type theory of
interest, reminiscent of Martin-Löf style predicative type systems. A key
design choice that impacts our proofs is the use of an untyped conversion
rule, inspired by Pure Type Systems~\citet{barendregt1991introduction}, and
specified through parallel reduction to normal form~\citet{barendregt}.
Because we believe that a good coverage of features makes our proof applicable
to a broad range of type systems, \lang{} incorporates some of the most common
features of dependent type theory, including an intensional identity type, a
boolean type, large elimination, dependent elimination, and a full universe
hierarchy. While our use of untyped conversion is nonstandard for MLTT, we
discuss how our proof can be adjusted for systems with a typed, judgmental
equality.

In Section~\ref{sec:logreldep}, we formulate the consistency property and
motivate where a proof by logical relations is required. The most important
contribution of this section is the definition of the logical relation as an
inductively defined relation and demonstrate how to reason about this
definition. We also show that the logical relation respects evaluation
(referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
prove that the logical relation is a partial function; that is, each type can
have at most one unique interpretation.

In Section~\ref{sec:logrelproof}, shows that our definitions can be the
subject of well-established reasoning techniques for logical relations. In
particular, we define a notion of semantic typing in terms of our logical
relation and prove the fundamental theorem, which states that syntactic typing
implies semantic typing. Consistency follows from this result as a corollary.

In Section~\ref{sec:extension}, extends the definition of the logical relation
to include terms with free variables, and proves that every well-typed term
has a $\beta$-normal form. After laying the foundation for handling type-level
computation in Section~\ref{sec:logreldep}, this extension requires mechanical
changes that directly mirror an analogous extension for a simply typed
language. We use this section to demonstrate the idea that once we have
established the base techniques, we can port ideas that have been studied in
the context of a simpler system to the dependently-typed setting.  We believe
it is possible to extend our logical relation to a Kripke-style and relational
model in a similar fashion.

In Section~\ref{sec:betaeta} we show how to incorporate the function $\eta$-law
into the conversion rule for \lang{}.

@@

First, the metatheoretic properties that we prove are indeed
simpler. Compared to Core Nuprl\citet{anand2014towards}, our system
lacks extensionality, which would require a relational model to
justify consistency. Compared to the systems from
\citet{decagda,nbeincoq, martin-lof-a-la-coq}, the conversion rule for
\lang{} is untyped and therefore we do not
need a Kripke-style relational model to prove $\Pi$-injectivity
among other properties. Furthermore, we prove the existence of normal
forms, which induces a simple \emph{normalize-and-compare}
procedure for type
conversion~\citet{pierce2004advanced}. \citet{nbeincoq, decagda}, on
the other hand, need
to show how their algorithmic conversion procedure is sound and complete
with respect to their respective declarative equational theory.
\scw{I'm getting confused by this paragraph. Does this reorganization sense:
  Our language is simpler than Nuprl, because it doesn't have extensional
  equality. It is simpler than Agda, because it doesn't have type-directed
  equality. Both of these cases require the definition of a binary logical
  relation, that defines a notion of semantic equality between terms. This
  relation justifies the injectivity of $\Pi$ types and justify the validity
  of $\eta$-conversion among other properties.}  \scw{Furthermore our proof is
  also simpler because we don't need prove the correctness of the NBE
  algorithm, which is used to show the decidability of Agda's type-directed
  equivalence. Therefore, we don't need to define this algorithm and show that
  it is sound and complete with respect to the type-directed
  equality. Instead, to show the decidability of our untyped equivalence, we
  need only show that terms have $\beta\eta$ normal forms. }
\yl{Makes sense. Though Abel's work doesn't use nbe but a recursive
  binary algorithm. Rewrote the paragraph above and commented out the original}

>>>>

Tutorial material on logical relations~\citep{skorstengaard2019introduction,
  harpertait, harperkripke, pierce2002types,
  pierce2004advanced,harper2016practical} is primarily focused on systems
with simple or polymorphic types. In that context, syntactic logical relations
can be defined as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation. However, neither of these techniques can be used to define a
logical relation in the context of a predicative dependent type theory, so a
young researcher might be excused for thinking that proofs that use logical
relations are not applicable for such languages.

But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs for dependent type theories}
Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code. As a result, their use of logical relations are
difficult to isolate from their surrounding contexts, and thus, not accessible
to casual readers.

@@

Not only does this connection aid readers that wish to, like us, adopt proof
assistants for their day-to-day use, we also find that this precision is
important for conveying the proof technique itself. Unlike properties that are
derivable through syntactic means, proofs by logical relation make demands on
the strength of the logic in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying
metatheory % may struggle to even
% convey the validity of their definition and
requires substantial
effort from the reader to even understand whether it is theoretically
possible to encode such proofs in a given ambient logic, such as those found
in modern proof assistants.

% The close correspondence between a typeset proof and a mechanized
% proof comes at the potential risk that the artifacts specific to
% theorem provers may become visible in the presented proof.
%However, for proofs by logical relation for
%dependent types, we find our approach advantageous for a few reasons.
Furthermore, while a proof assistant may reject perfectly valid definitions,
it helps us keep our definition precise and unambiguous. A precise definition
is crucial as the reasoning principles we are allowed to use are directly tied
to the definition, which is sometimes obscured by concise notations that
diverge too far from a formal proof.

@@

In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type theory of
interest, reminiscent of Martin-Löf style predicative type systems. A key
design choice that impacts our proofs is the use of an untyped conversion
rule, inspired by Pure Type Systems~\citep{barendregt1991introduction}, and
specified through parallel
reduction~\citep{takahashi-parallel-reduction, barendregt:lambda-calculi-with-types} % to normal form~\citet{barendregt}
.
Because we believe that a good coverage of features makes our proof applicable
to a broad range of type systems, \lang{} incorporates some of the most common
features of dependent type theory, including an intensional identity type, a
boolean type, large elimination, dependent elimination, and a full universe
hierarchy. While our use of untyped conversion is nonstandard for MLTT, we
discuss how our proof can be adjusted for systems with a typed, judgmental
equality.

In Section~\ref{sec:logreldep}, we formulate the consistency property and
motivate where a proof by logical relations is required. The most important
contribution of this section is the definition of the logical relation as an
inductively defined relation and demonstration of how to reason about this
definition. We also show that the logical relation respects evaluation
(referred to as the irrelevance property in \citet{martin-lof-a-la-coq}), and
prove that the logical relation is a partial function; that is, each type can
have at most one unique interpretation.

In Section~\ref{sec:logrelproof}, we show that our definitions can be the
subject of well-established reasoning techniques for logical relations. In
particular, we define a notion of semantic typing in terms of our logical
relation and prove the fundamental theorem, which states that syntactic typing
implies semantic typing. Consistency follows from this result as a corollary.

In Section~\ref{sec:extension}, we extend the definition of the logical relation
to include terms with free variables, and prove that every well-typed term
has a $\beta$-normal form. After laying the foundation for handling type-level
computation in Section~\ref{sec:logreldep}, we show how this extension requires mechanical
changes that directly mirror an analogous extension for a simply typed
language. We use this section to demonstrate the idea that once we have
established the base techniques, we can port ideas that have been studied in
the context of a simpler system to the dependently-typed setting.  We believe
it is possible to extend our logical relation to a Kripke-style and relational
model in a similar fashion.

In Section~\ref{sec:betaeta}, we show how to incorporate the function $\eta$-law
into the conversion rule for \lang{}.

@@

First, the metatheoretic properties that we prove are indeed
simpler. Compared to Core Nuprl~\citep{anand2014towards}, our system
lacks extensionality, which would require a relational model to
justify consistency. Compared to the systems from
\citet{decagda,nbeincoq, martin-lof-a-la-coq}, the conversion rule for
\lang{} is untyped and therefore we do not
need a Kripke-style relational model to prove $\Pi$-injectivity
among other properties. Furthermore, we prove the existence of normal
forms, which induces a simple \emph{normalize-and-compare}
procedure for type
conversion~\citet{pierce2004advanced}. \citet{nbeincoq, decagda}, on
the other hand, need
to show how their algorithmic conversion procedure is sound and complete
with respect to their respective declarative equational theory.
\scw{I'm getting confused by this paragraph. Does this reorganization sense:
  Our language is simpler than Nuprl, because it doesn't have extensional
  equality. It is simpler than Agda, because it doesn't have type-directed
  equality. Both of these cases require the definition of a binary logical
  relation, that defines a notion of semantic equality between terms. This
  relation justifies the injectivity of $\Pi$ types and justify the validity
  of $\eta$-conversion among other properties.}  \scw{Furthermore our proof is
  also simpler because we don't need prove the correctness of the NBE
  algorithm, which is used to show the decidability of Agda's type-directed
  equivalence. Therefore, we don't need to define this algorithm and show that
  it is sound and complete with respect to the type-directed
  equality. Instead, to show the decidability of our untyped equivalence, we
  need only show that terms have $\beta\eta$ normal forms. }
\yl{Makes sense. Though Abel's work doesn't use nbe but a recursive
  binary algorithm. Rewrote the paragraph above and commented out the original}
